Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-06 submission/manuscript.Rmd
add motif to text

- read in the homer results txt file above here

- Germline genes repressed by other histone modifying enzymes have shared genomic features around the TSS to suppress their expression, such as motifs for E2F6/DP1 (E2F) and MGA/MAX (E-box) binding.
	- We asessed genomic features enriched at KDM5C-bound versus unbound gemrline gene promoters using HOMER.
		- window I analyzed

We then compared KDM5C binding at promoters of RNA-seq DEGs to determine if the germline mRNAs de-repressed in _Kdm5c_-KO EpiLCs and brain are direct targets of KDM5C (Figure 4F). In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEGs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%). Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs (Figure 4H). Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above. Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 4F). In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS (Figure 4G). Altogether, this suggests KDM5C decommssions germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters.

@@

We then evaluated KDM5C's role in germline gene silencing during ESC to EpiLC differentiation. We first characterized KDM5C's mRNA and protein expression at 0, 24, and 48 hours of differentiation (Figure 5C). While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 5D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 5E). To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs). Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 5F).

@@

>>>>

- read in the homer results txt file above here
- Add in euler KDM5C-bound vs KDM5C-unbound coloring
		- Make sure instances window is 500 bp, change if not

- Germline genes repressed by other histone modifying enzymes have shared genomic features around the TSS to suppress their expression, such as motifs for E2F6/DP1 (E2F) and MGA/MAX (E-box) and CpG islands (CGIs). We asessed transcription factor motifs enriched at germline gene promoters that are either bound or unbound by KDM5C in EpiLCs using HOMER[XXX] (TSS +/- 500 bp, q-value < 0.1). MAX and E2F6 binding sites were significantly enriched at KDM5C-bound germline genes, with about one third of KDM5C-bound promoters contained one or both of these motifs (Figure 4). Contrastingly, KDM5C-unbound germline genes were enriched for RFX transcription factor motifs, a family of transcription factors important for ciliogenesis. The RFX X-box motif is present at XXX% of KDM5C-unbound promoters (Figure 4).

	- Enriched RFX transcription factors included the spermiogenesis regulator RFX2, whose mRNA is also significantly upregulated in _Kdm5c_-KO EpiLCs (Figure 4). 
	 - https://www.nature.com/articles/srep20435
	 - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4498915/
	 - https://pubmed.ncbi.nlm.nih.gov/32725242/

Finally, we compared KDM5C binding at promoters of RNA-seq DEGs to determine if the germline mRNAs de-repressed in _Kdm5c_-KO EpiLCs and brain are direct targets of KDM5C (Figure 4F). In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEGs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%). Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs (Figure 4H). Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above. Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 4F). In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS (Figure 4G). Altogether, this suggests KDM5C decommssions germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters.

@@

We then evaluated KDM5C's role in germline gene silencing during ESC to EpiLC differentiation. We first characterized KDM5C's mRNA and protein expression at 0, 24, and 48 hours of differentiation (Figure 5C). While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 5D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 5E).

- Germline genes are known to accumulate CpGme at (CGIs).
To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs). Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 5F).

- discussion 
 - 	- However, CPG islands greatly detrmine KDM5C recruitment - KDM5C is known to be enriched at CGIs.)

@@

Motif Discussion
- only 1/3 had max/e2f - Unlikely to be the major recruitment mechanism for KDM5C (discussion?)
